A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-04 tex/codatadp.tex
more

Datatypes and codatatypes can be characterized axiomatically. They share basic
properties of constructors and selectors. All properties below are implicitly
universally quantified and range over all
$i$, $j$, $\jPrime$, and $k$ within bounds:
% and to all possible splits of the
%$n$-ary constructor $\const{C}_{i\negvthinspace j}$'s argument list into $\bar x,
%y, \bar z$:
%
\[
\begin{aligned}[t]
\text{Distinctness:}\quad
  & %\forall \bar x, \bar y.\;\,
    \const{C}_{i\negvthinspace j}(\bar x) \tneq \const{C}_{i\negvthinspace \jPrime}(\bar y) \qquad\text{if $j \not= \jPrime$}
  \\
\text{Injectivity:}\quad
  & %\forall x_1,\ldots,x_{n_{i\negvthinspace j}} y.\;\,
    \const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}}) \teq \const{C}_{i\negvthinspace j}(x_1,\ldots,x_{k-1},y,x_{k+1},\ldots,x_{n_{i\negvthinspace j}}) \longrightarrow x_k \teq y
  \\
\text{Exhaustiveness:}\quad
  & \is{C}_{i1}(x) \mathrel\lor \cdots \mathrel\lor \is{C}_{im_i}(x)
  \\
\text{Selection:}\quad
  & s_{ij}^k(\const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}})) = x_k
\end{aligned}
\]
%
Expressed in the algebraic jargon, exhaustiveness ensures that ``no junk''
exists, whereas distinctness and injectivity ensure that ``no confusion''
arises.

Datatypes are additionally characterized by an induction axiom schema:
%
\[
\begin{aligned}[t]
\text{Induction:}\quad
\AXC{$\forall x_1 \ldots \vthinspace x_{n_{ij}}.\; \bigl(\bigwedge_{\,k\,} \mathit{IH}_{ij}^k[\bar x_k]\bigr) \longrightarrow P_i[C_{ij}(\bar x)]$\quad\text{for each $i,j$}}
\UIC{$P_1[v_1] \mathrel\land \cdots \mathrel\land P_l[v_l]$}
\DP
\end{aligned}
\]
where the induction hypothesis $\mathit{IH}_{ij}^k(x)$
denotes either $P_{\iPrime}(x)$ if there exists some $i$ such that this is type
correct or else $\top$.
%
The axiom schema ensures that we get a smallest model. For example,
for a datatype of natural numbers constructed from $\const{Z}$ and $\const{S}$,
this rules out nonstandard models, which could contain cyclic values (e.g.,
an $n$ such that $n \teq \const{S}(n)$) or even infinite acyclic values
$\const{S}(\const{S}(\ldots))$.

For codatatype, the dual notion is that of a coinduction principle that
depends on witnesses $R$ that are required to be bisimulations:
%
\[
\begin{aligned}[t]
\text{Conduction:}\quad
\AXC{for each i: $R_i[x_i, y_i]$}
\AXC{for each i: $(\forall x y. R_i[x, y] \longrightarrow \bigwedge_k s_{ij}^k(x) \sim s_{ij}^k(y))$}
\BIC{$x_1 \teq y_1 \& ... \& x_l \teq y_l$}
\DP
\end{aligned}
\]

@@

  * Strategies
  * Implementation
    * mention that discriminators are their own function symbols
    * what is an SMT theory solver
    * cooperation with Nelson-Oppen / integration into SMT solver
    * "mu" stuff
    * no "internal search"
  * Enumerator?
    * for model generation

>>>>

Datatypes and codatatypes can be characterized axiomatically. They share basic
properties of constructors and selectors. All properties below are implicitly
universally quantified and range over all
$i$, $j$, $\jPrime$, and $k$ within bounds:
% and to all possible splits of the
%$n$-ary constructor $\const{C}_{i\negvthinspace j}$'s argument list into $\bar x,
%y, \bar z$:
%
\[
\begin{aligned}[t]
\text{Distinctness:}\quad
  & %\forall \bar x, \bar y.\;\,
    \const{C}_{i\negvthinspace j}(\bar x) \tneq \const{C}_{i\negvthinspace \jPrime}(\bar y) \qquad\text{if $j \not= \jPrime$}
  \\
\text{Injectivity:}\quad
  & %\forall x_1,\ldots,x_{n_{i\negvthinspace j}} y.\;\,
    \const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}}) \teq \const{C}_{i\negvthinspace j}(x_1,\ldots,x_{k-1},y,x_{k+1},\ldots,x_{n_{i\negvthinspace j}}) \longrightarrow x_k \teq y
  \\
\text{Exhaustiveness:}\quad
  & \is{C}_{i1}(x) \mathrel\lor \cdots \mathrel\lor \is{C}_{im_i}(x)
  \\
\text{Selection:}\quad
  & s_{i\negvthinspace j}^k(\const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}})) = x_k
\end{aligned}
\]
%
Expressed in the algebraic jargon, exhaustiveness ensures that ``no junk''
exists, whereas distinctness and injectivity ensure that ``no confusion''
arises.

Datatypes are additionally characterized by an induction axiom schema:
%
\[
\begin{aligned}[t]
\text{Induction:}\quad
\AXC{$\forall x_1 \ldots \vthinspace x_{n_{i\negvthinspace j}}.\; \bigl(\bigwedge_{\,k\,} \mathit{IH}_{i\negvthinspace j}^k[\bar x_k]\bigr) \longrightarrow P_i[C_{i\negvthinspace j}(\bar x)]$\quad\text{for each $i,j$}}
\UIC{$P_1[v_1] \mathrel\land \cdots \mathrel\land P_l[v_l]$}
\DP
\end{aligned}
\]
where the induction hypothesis $\mathit{IH}_{i\negvthinspace j}^k(x)$
denotes either $P_{\iPrime}(x)$ if there exists some $i$ such that this is type
correct or else $\top$.
%
The axiom schema ensures that we get a smallest model. For example,
for a datatype of natural numbers constructed from $\const{Z}$ and $\const{S}$,
this rules out nonstandard models, which could contain cyclic values (e.g.,
an $n$ such that $n \teq \const{S}(n)$) or even infinite acyclic values
$\const{S}(\const{S}(\ldots))$.

For codatatype, the dual notion is that of a coinduction principle that
depends on witnesses $R$ that are required to be bisimulations:
%
\[
\begin{aligned}[t]
\text{Conduction:}\quad
\AXC{for each i: $R_i[x_i, y_i]$}
\AXC{for each i: $(\forall x y. R_i[x, y] \longrightarrow \bigwedge_k s_{i\negvthinspace j}^k(x) \sim s_{i\negvthinspace j}^k(y))$}
\BIC{$x_1 \teq y_1 \& ... \& x_l \teq y_l$}
\DP
\end{aligned}
\]

@@

  * Strategies
  * Implementation
    * mention that discriminators are their own function symbols
    * what is an SMT theory solver
    * sharing is caring (and other "refinements" on calculus)
    * cooperation with Nelson-Oppen / integration into SMT solver
    * "mu" stuff
    * no "internal search"
  * Enumerator?
    * for model generation
